---
title: 'CHERISH: Collaborative for Hospitalised Elders Reducing the Impact of Stays
  in Hospital. Analysis of 30-day outcomes.'
author: "Adrian Barnett"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  word_document:
    toc: yes
    toc_depth: 2
    reference_docx: rmarkdown-styles-reference.docx
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE, comment='', dpi=400)
options(width=1000) # Wide pages
options(scipen=999) # avoid scientific presentation
library(stringr)
library(R2WinBUGS)
library(dplyr)
library(janitor)
library(flextable)
library(pander)
#panderOptions('keep.trailing.zeros', TRUE) # causes weird error with summarytools
panderOptions('table.emphasize.rownames', FALSE)
panderOptions('table.split.table', Inf)
panderOptions('table.split.cells', Inf)
panderOptions('big.mark', ',')
library(broom)
#library(mitools) # for combining estimates from imputed data
library(ordinal)
library(summarytools)
# global options for summary tools
st_options(plain.ascii = FALSE,       # Always use this option in Rmd documents
            style = "rmarkdown",        # This too
            round.digits = 0, 
            headings = FALSE, 
            footnote = NA,             # Avoids footnotes which would clutter the results
            subtitle.emphasis = TRUE  # Improves layout with some rmarkdown themes
) 
# graphics stuff
library(ggplot2)
g.theme = theme_bw() + theme(panel.grid.minor = element_blank())
cbPalette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# 
source('99_functions.R')
## main data:
load('data/Analysis.Ready30.RData') # from MakeDataFeb2021.R
load('data/MICE.Results.impute.RData') # from impute.mice.R
# merge imputed data
data = dplyr::select(data, -iadl_baseline, -starts_with('spmsq')) %>% # remove missing variables
    left_join(impute[[1]], by='subject_num') %>%
  clean_names()

# minor edits
data  = mutate(data, 
               age_c = (age-75)/5, # scale age
               spmsq_score_admission_c = (spmsq_score_admission - 8)/(-3), # scale and also change direction
               cci_unadjusted_for_age_c = (cci_unadjusted_for_age - 2)/2, #
               # reverse order so that death is worst category
               status_d30 = forcats::fct_rev(status_d30),
               status_d30_adl = forcats::fct_rev(status_d30_adl),
               status_d30_iadl = forcats::fct_rev(status_d30_iadl)) %>%
  filter(source01 == 'Post-intervention') # just post-intervention

# make binary baseline variables and gender variable for winbugs
data = mutate(data, 
              female = as.numeric(gender=='Female'), # numeric binary variables
              baseline_a = as.numeric(adl_baseline>0),
              baseline_i = as.numeric(iadl_baseline>0))

# scramble intervention (turn off when ready)
scramble = TRUE # 
```

# Scrambled intervention group

This document contains the statistical analyses for the 30-day outcomes of the CHERISH study. The initial document was produced using a scrambled intervention group by randomly assigning participants to intervention or control wards. This allowed us to finalise the statistical analyses plan and ensure that all investigators understood the analyses before the real intervention was used.

```{r scramble, include=FALSE}
# scramble from now on
if(scramble==TRUE){
  set.seed(123456)
  data$intervention = sample(data$intervention, size=nrow(data), replace=FALSE)
  data$intervention_factor = sample(data$intervention_factor, size=nrow(data), replace=FALSE)
  data$source01 = sample(data$source01, size=nrow(data), replace=FALSE)
}
```

# Summary tables

We only use the post-intervention data.

### Functional measures at 30 days 

#### ADL

```{r, results='asis'}
with(data, ctable(adl_total_d30, intervention_factor, dnn=c(' ',''), prop='c', round.digits=0, useNA='ifany'))
```

#### IADL

```{r, results='asis'}
with(data, ctable(iadl_total_d30, intervention_factor, dnn=c(' ',''), prop='c', round.digits=0, useNA='ifany'))
```

### 4-level composite 

```{r, results='asis'}
with(data, ctable(status_d30, intervention_factor, dnn=c(' ',''), prop='c', round.digits=0, useNA='ifany'))
```


### 4-level composite using ADL

```{r, results='asis'}
with(data, ctable(status_d30_adl, intervention_factor, dnn=c(' ',''), prop='c', round.digits=0, useNA='ifany'))
```

### 4-level composite using IADL

```{r, results='asis'}
with(data, ctable(status_d30_iadl, intervention_factor, dnn=c(' ',''), prop='c', round.digits=0, useNA='ifany'))
```

# Ordinal models

We used an ordinal regression model to examine the ordered four categories of: 

* 1. Dead
* 2. Still in hospital
* 3. Out of hospital level 1
* 4. Out of hospital level 2

The definition for categories 3 and 4 depends on the outcome. 

The model examines the odds of being in a worse (higher) category [@Dobson18]. A logistic regression model examines the odds of one category compared to another, such as dead versus alive. In this model we have four categories and examine the odds of being in a worse category. In other words, for whatever category a patient is currently in, what are their odds of worsening.

We adjusted for age and gender in the ordinal model. We also adjusted for baseline status, so for the ADL at 30 days model, we adjusted for baseline ADL. We also adjust for Charlson comorbidity score, admissions SPMSQ score and elective status. We also included a random intercept for each hospital to adjust for the potential correlation in results from patients from the same hospital.

The model was fitted using a Bayesian approach.

### Ordinal model for ADL

For this model categories 3 and 4 are:

* 3. Out of hospital but poorer ADL function
* 4. Out of hospital recovered ADL function

##### Odds ratios

```{r, include=FALSE}
# keep code, but do not show results
omodel =  clm(status_d30_adl ~ age_c + gender + I(adl_baseline>0) + intervention_factor + cci_unadjusted_for_age_c + spmsq_score_admission_c + elective, data=data)
res = tidy(omodel, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(p.value = format.pval(p.value, eps=0.0001, digits=4)) %>%
  filter(coef.type == 'location') %>% # just regression parameters
  mutate(term = nice_rename(term, scramble=scramble),
         estimate = roundz(estimate, 2),
         CI = paste(roundz(conf.low, 2), ' to ', roundz(conf.high,2), sep='')
         #p.value = format.pval(p.value, eps=0.0001, digits=4)
         ) %>%
  select(term, estimate, CI, p.value) 
#pander(res, style='rmarkdown') # needed rmarkdown style
```

```{r, include=FALSE}
# for text below
age.odds = res$estimate[grep('Age', res$term)]
# get fitted values
pred <- predict(omodel, newdata = subset(data, select = -status_d30_adl))$fit
class <- predict(omodel, newdata = subset(data, select = -status_d30_adl), type='class')$fit
```

```{r}
## using bayes for random intercept
## two versions, one with and one without hospital
# a) with hospital
bugs.model = 'bugs_latent.txt'
# create external text file with bugs model
bugs = file(bugs.model, 'w')
cat('model{
## Likelihood
for (i in 1:N){
  freq[i, 1:ncat] ~ dmulti(pi[i, 1:ncat], 1);
# Cumulative probability of > category k given cutpoint
  for (k in 1:ncat-1){
    logit(Q[i,k]) <- inprod(X[i,1:n.beta], beta[1:n.beta]) + gamma.c[hosp[i]]-C[k];
  }
# Calculate probabilities
  pi[i,1] <- 1 - Q[i,1]; # Pr(cat=1)=1-Pr(cat>1);
  for (k in 2:(ncat-1)) {
    pi[i,k] <- Q[i,k-1]-Q[i,k]; # Pr(cat>k-1)-Pr(cat>k);
  }
  pi[i,ncat] <- Q[i,ncat-1]; # Pr(cat=k)=Pr(cat>k-1);
}
## Priors
for (k in 1:n.beta){beta[k] ~ dnorm(0,0.00001);}
for (k in 1:4){
  gamma[k] ~ dnorm(0, tau.gamma);
  gamma.c[k] <- gamma[k] -mean.gamma
} # random hospital effects
mean.gamma <- mean(gamma[1:4])
tau.gamma ~ dgamma(1,1)
# ordered cut-offs
C[1] <- 0; # for identifiability
C[2] ~ dunif(C[1],C[3]);
C[3] ~ dunif(C[2],10);
}\n', file=bugs)
close(bugs)

# b) without hospital
bugs.model = 'bugs_latent_no_hospital.txt'
# create external text file with bugs model
bugs = file(bugs.model, 'w')
cat('model{
## Likelihood
for (i in 1:N){
  freq[i, 1:ncat] ~ dmulti(pi[i, 1:ncat], 1);
# Cumulative probability of > category k given cutpoint
  for (k in 1:ncat-1){
    logit(Q[i,k]) <- inprod(X[i,1:n.beta], beta[1:n.beta]) - C[k];
  }
# Calculate probabilities
  pi[i,1] <- 1 - Q[i,1]; # Pr(cat=1)=1-Pr(cat>1);
  for (k in 2:(ncat-1)) {
    pi[i,k] <- Q[i,k-1]-Q[i,k]; # Pr(cat>k-1)-Pr(cat>k);
  }
  pi[i,ncat] <- Q[i,ncat-1]; # Pr(cat=k)=Pr(cat>k-1);
}
## Priors
for (k in 1:n.beta){beta[k] ~ dnorm(0,0.00001);}
# ordered cut-offs
C[1] <- 0; # for identifiability
C[2] ~ dunif(C[1],C[3]);
C[3] ~ dunif(C[2],10);
}\n', file=bugs)
close(bugs)

# run the ADL model
dep_vars = vars = c('age_c','female','baseline_a','cci_unadjusted_for_age_c','spmsq_score_admission_c','elective','intervention','hospital') # list of dependent variables
adl_ests = run_winbugs(indata=data, vars=dep_vars, outcome='ADL')
adl_ests$dic$model = 'Full model'
# for text below
age.odds = adl_ests$ests$mean[grep('Age', adl_ests$ests$Variable)]
# nice table
ests = flextable(adl_ests$ests)
ests = autofit(ests)
ests
```

The chains converged well and the largest R-hat was `r roundz(adl_ests$max_rhat, digits=2)`.

The estimates are odds ratios (and 95% credible intervals) for the outcome of being in a worse category. So each 5 year increase in age increases the odds of being a worse category by `r age.odds`.



### Ordinal model for IADL

For this model categories 3 and 4 are:

* 3. Out of hospital but poorer IADL function
* 4. Out of hospital recovered IADL function

##### Odds ratios

```{r, include=FALSE}
# keep code but use Bayesian model below instead
omodel =  clm(status_d30_iadl ~ age_c + gender + I(iadl_baseline>0) + intervention_factor + cci_unadjusted_for_age_c + spmsq_score_admission_c + elective, data=data)
res = tidy(omodel, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(p.value = format.pval(p.value, eps=0.0001, digits=4)) %>%
  filter(coef.type == 'location') %>% # just regression parameters
  mutate(term = nice_rename(term, scramble=scramble),
         estimate = roundz(estimate, 2),
         CI = paste(roundz(conf.low, 2), ' to ', roundz(conf.high,2), sep='')
         #p.value = format.pval(p.value, eps=0.0001, digits=4)
         ) %>%
  select(term, estimate, CI, p.value) 
pander(res, style='rmarkdown') # needed rmarkdown style
```

```{r}
# as above model, but with IADL as outcome and baseline
dep_vars = vars = c('age_c','female','baseline_i','cci_unadjusted_for_age_c','spmsq_score_admission_c','elective','intervention','hospital') # list of dependent variables
iadl_ests = run_winbugs(indata=data, vars=dep_vars, outcome='IADL')
iadl_ests$dic$model = 'Full model'
# make nice table
nice_table = flextable(iadl_ests$ests)
nice_table = autofit(nice_table)
nice_table
```

The chains converged well and the largest R-hat was `r roundz(iadl_ests$max_rhat, digits=2)`.

# Is there some colinearity with elective surgery and ward?

There may be a strong overlap with elective surgery and ward, increasing the risk of colinearity.
The cross-tabulation of elective surgery by ward is below, showing the counts and row percents.

```{r, results='asis'}
data = mutate(data, elective_nice = factor(elective, levels=0:1, labels=c('No','Yes')))
with(data, ctable(ward, elective_nice, dnn=c(' ',''), prop='r', round.digits=0, useNA='ifany'))
```

The table shows there are five wards that 5 or fewer elective patients. 

To examine if there is colinearity, we can look at the change in estimates if we remove ward or elective surgery.

#### Estimates without hospital (ADL)

```{r}
dep_vars = vars = c('age_c','female','baseline_a','cci_unadjusted_for_age_c','spmsq_score_admission_c','elective','intervention') # list of dependent variables
adl_ests_no_hospital = run_winbugs(indata=data, vars=dep_vars, outcome='ADL')
adl_ests_no_hospital$dic$model = 'Without hospital'
# make nice table
nice_table = flextable(adl_ests_no_hospital$ests)
nice_table = autofit(nice_table)
nice_table
```

#### Estimates without elective (ADL)

```{r}
dep_vars = vars = c('age_c','female','baseline_a','cci_unadjusted_for_age_c','spmsq_score_admission_c','hospital','intervention') # list of dependent variables
adl_ests_no_elective = run_winbugs(indata=data, vars=dep_vars, outcome='ADL')
adl_ests_no_elective$dic$model = 'Without elective'
# make nice table
nice_table = flextable(adl_ests_no_elective$ests)
nice_table = autofit(nice_table)
nice_table
```

#### Estimates without hospital (IADL)

```{r}
dep_vars = vars = c('age_c','female','baseline_i','cci_unadjusted_for_age_c','spmsq_score_admission_c','elective','intervention') # list of dependent variables
iadl_ests_no_hospital = run_winbugs(indata=data, vars=dep_vars, outcome='IADL')
iadl_ests_no_hospital$dic$model = 'Without hospital'
# make nice table
nice_table = flextable(iadl_ests_no_hospital$ests)
nice_table = autofit(nice_table)
nice_table
```


#### Estimates without elective (IADL)

```{r}
dep_vars = vars = c('age_c','female','baseline_i','cci_unadjusted_for_age_c','spmsq_score_admission_c','hospital','intervention') # list of dependent variables
iadl_ests_no_elective = run_winbugs(indata=data, vars=dep_vars, outcome='IADL')
iadl_ests_no_elective$dic$model = 'Without elective'
# make nice table
nice_table = flextable(iadl_ests_no_elective$ests)
nice_table = autofit(nice_table)
nice_table
```

### Plot of estimates

Here we compare the estimates for the three models.

```{r, fig.width=8}
## prepare data
# IADL
iadl_ests$ests$outcome = 'IADL'
iadl_ests$ests$model = 'Full model'
iadl_ests_no_elective$ests$outcome = 'IADL'
iadl_ests_no_elective$ests$model = 'No elective'
iadl_ests_no_hospital$ests$outcome = 'IADL'
iadl_ests_no_hospital$ests$model = 'No hospital'
# ADL
adl_ests$ests$outcome = 'ADL'
adl_ests$ests$model = 'Full model'
adl_ests_no_elective$ests$outcome = 'ADL'
adl_ests_no_elective$ests$model = 'No elective'
adl_ests_no_hospital$ests$outcome = 'ADL'
adl_ests_no_hospital$ests$model = 'No hospital'
# concatenate
to_plot = bind_rows(iadl_ests$ests,
                    iadl_ests_no_hospital$ests,
                    iadl_ests_no_elective$ests,
                    adl_ests$ests,
                    adl_ests_no_hospital$ests,
                    adl_ests_no_elective$ests) %>%
  mutate(mean = as.numeric(mean))
compare_plot = ggplot(to_plot, aes(x=variable, y=mean, col=model, shape=model))+
  geom_point(size=4)+
  scale_color_manual(NULL, values=c("#FF7F24", "#B4EEB4", "#1C86EE"))+
  scale_shape_manual(NULL, values=0:2)+
  xlab('')+
  geom_hline(lty=2, yintercept=1)+
  theme_bw()+
  coord_flip()+
  facet_wrap(~outcome)
compare_plot
```

##### Deviance information criterion

We can compare the fit of the model using the deviance information criterion. The lower the DIC, the better the model. "pD" is the approximate number of parameters used by the model, which does not have to be an integer in a Bayesian model. 

```{r}
# combine DIC estimates
DIC = bind_rows(adl_ests$dic,
                adl_ests_no_elective$dic,
                adl_ests_no_hospital$dic,
                iadl_ests$dic,
                iadl_ests_no_elective$dic,
                iadl_ests_no_hospital$dic) %>%
  select('model','outcome','pD',"DIC") %>%
  mutate(pD = roundz(pD, 1),
         DIC = roundz(DIC, 1))
DIC = flextable(DIC)
DIC = autofit(DIC)
DIC
```

The table shows the three alternative models for ADL and IADL. The best model for ADL is the full model with both elective and hospital. The best for for IADL is without elective, although it is not clearly better than the full model, but could be preferred on the grounds of parsimony. 

### Software

This report and the statistical analyses were made using Rmarkdown with `r version$version.string`.

### References
